ST563 Final Project

### 2021-04-24 DRAFT.Rmd
Added our names to the front page. Fixed a couple small bugs and made minor wording revisions.

---
title: "ST563 Final Project"
author: "Final Project Group 6"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    toc: true
    toc_depth: 1
fontsize: 12pt
---

@@

# Classification Models
## Classification Tree
The basic classification tree is based on partitioning the data into subgroups using simple binary splitting. Initially, all objects are considered a single group. Then, the group is repeatedly split into two subgroups based on the criteria of a certain variable. In the classification setting, an observation is classified in a specific region with majority vote. Trees are grown as big as possible and then pruned back using cost-complexity pruning. This is done to ensure the data is not being overfit. The trade-off here is that pruning increases the bias.

@@

# Regression
## Multiple Linear Regression
Multiple Linear Regression is the extension of simple linear regression to a set of multiple predictor variables. The goal of multiple linear regression is to model the linear relationship between a continuous response and two or more predictor variables. Like simple linear regression, coefficient estimates are found by minimizing the sum of the squared errors. The formula for a multiple linear regression model with $p$ predictor variables is as follows:

$$
\begin{aligned}
&Y_i = \beta_0 + \beta_1X_{i1} + \beta_2X_{i2} + ... + \beta_pX_{ip} + \epsilon_i, \\
& \quad \text{where observation } i = 1, ..., n \\
& \quad Y_i = \text{dependent variable (response)} \\
& \quad X_{ij} = \text{independent variables for } j = 1, ... , p \\
& \quad \beta_0 = \text{y-intercept} \\
& \quad \beta_j = \text{slope coefficients for each variable} \\
& \quad \epsilon_i = \text{residuals (error term of model)} \\
\end{aligned}
$$
Although it is possible to use every predictor in the MLR model, Best Subsets Selection selected the variables that best explain the variation in the response. The `leaps` library and the `regsubsets()` function in R produces comparisons of all possible models from the given set of predictors. Best Subset Selection compares all the possible combinations of predictors to select the best model that produces the smallest test error. Three different model criteria were evaluated to compare models of different amounts of predictor variables: Mallow's Cp, BIC, and $R^2_{\text{adj}}$.

@@

Although each of the three model criteria suggested a different number of predictor variables, they were all within the same region (between 6 and 8 predictors). All three produced a similar error rate, so the theory of Occam's Razor was implemented and the simplest, six-variable model was selected. This was the model favored by BIC, which gives the greatest penalty for complexity in models. The six variables used in the regression context are then `volatile.acidity`, `chlorides`, `total.sulfur.dioxide`, `pH`, `sulphates`, and `alcohol`. All of the variables from the classification selection appear with the exception of density. In the regression context, `pH` was the variable that proved to be a rather unintuitive inclusion, because it is similar to density in that it is not easily detected by human senses. The inclusion of these two variables in the model portray the importance of the chemical properties of a bottle of wine when it comes to quality.

@@

# Obtain RMSE and results for linear model
lm.reg.trn.pred <- predict(lm.reg, newdata = trn)
lm.reg.trn.rmse <- sqrt(mean((lm.reg.trn.pred - tst$quality)^2))

@@

Residuals indicate that the MLR assumptions are mostly satisfied. The residuals vs. fitted values plot (top left) has unusual stratified lines, but this is caused by the observed responses being integer values of 4, 5, 6, 7, etc. rather than continuous values. The plot does not show heteroscedasticity and residuals are spread evenly around 0 over the entire range of fitted values. When residuals are standardized, as in the top right plot, one would expect about 95% to be inside the [-1.96, 1.96] interval; the actual proportion is `r round(scale.res.2std, 3)`. The QQ-plot in the bottom left compares the normality of residuals against a theoretical distribution. Except for a slight downward bias of the lowest 5% of residuals at 2 standard deviations below 0, the data fits a normal distribution almost perfectly. On the bottom right plot, high leverage points (those with leverage $\ge$ 0.03) are not dramatically biased, being spread both above and below the 0 residual. Cook's Distance tops out at about 0.05, well below the generally accepted threshold of 1 that is considered highly influential. These plots taken together suggest that the data generally fulfills the requirements for modeling using MLR.

The largest benefit to using a multiple linear regression model is its interpretability and ease of use. The coefficients can be computed efficiently and they have clear and easy interpretations. It is also possible to make inference on any of the variables or linear combination of variables. The most significant drawback of using the multiple linear regression model is that it is subject to distributional assumptions, meaning that if the assumptions are violated, the estimates and inference will be unstable at best and invalid at worst. Based on the residual plots, though, it appears that the model does not violate any of the model assumptions.

@@

Based on cross validation, the random forest model for regression using two randomly-selected variables considered at each split was selected. This model produced a test error rate of `rf.reg.test.err`.

@@

In the regression setting, the values of root mean squared error (RMSE) for both models were relatively low. The RMSE for the random forest was `r df.reg[2,2]` and the RMSE for the multiple linear regression model was `r df.reg[1,2]`, further supporting the hypothesis that the random forest would outperform the other models.

@@

Based on the results of the study, prediction of if a given wine has high or low quality was achieved with about 70-80% accuracy. Despite not being able to achieve 80% prediction accuracy with the basic classification tree, the random forest model showed an improvement and predicted with a higher accuracy. Although this is a *good* result considering the non-information rate of 53%, it is by no means excellent. These results indicate that machine learning algorithms could be a good technique when trying to predict wine quality, but that there is room for improvement. With a more formal, industry-expert-guided variable selection process, the team believes that higher prediction accuracy can be achieved. Machine learning models could potentially be combined with other techniques to add value to winemakers.

>>>>

---
title: "ST563 Final Project"
author:
  - Final Project Group 6
  - Mana Azizsoltani, Rong Huang, Yun Ji, Jackie Steffan
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    toc: true
    toc_depth: 1
fontsize: 12pt
---

@@

# Classification Models
## Classification Tree
The basic classification tree is based on partitioning the data into subgroups using simple binary splitting. Initially, all observations are considered a single group. Then, the group is repeatedly split into two subgroups based on the criteria of a certain variable. In the classification setting, an observation is classified in a specific region with majority vote. Trees are grown as big as possible and then pruned back using cost-complexity pruning. This is done to ensure the data is not being overfit. The trade-off here is that pruning increases the bias.

@@

# Regression
## Multiple Linear Regression
Multiple linear regression is the extension of simple linear regression to a set of multiple predictor variables. The goal of multiple linear regression is to model the linear relationship between a continuous response and two or more predictor variables. Like simple linear regression, coefficient estimates are found by minimizing the sum of the squared errors. The formula for a multiple linear regression model with $p$ predictor variables is as follows:

$$
\begin{aligned}
&Y_i = \beta_0 + \beta_1X_{i1} + \beta_2X_{i2} + ... + \beta_pX_{ip} + \epsilon_i, \\
& \quad \text{where observation } i = 1, ..., n \\
& \quad Y_i = \text{dependent variable (response)} \\
& \quad X_{ij} = \text{independent variables for } j = 1, ... , p \\
& \quad \beta_0 = \text{y-intercept} \\
& \quad \beta_j = \text{slope coefficients for each variable} \\
& \quad \epsilon_i = \text{residuals (error term of model)} \\
\end{aligned}
$$
Although it is possible to use every predictor in the MLR model, the method of Best Subsets Selection can help choose the predictors that best explain the variation in the response, for any given model containing a fixed number of predictors. The `leaps` library and the `regsubsets()` function in R produces comparisons of all possible models from the given set of predictors. Best Subset Selection compares all the possible combinations of predictors to select the best model that produces the smallest test error. Three different model criteria were evaluated to compare models of different numbers of predictor variables: Mallow's Cp, BIC, and $R^2_{\text{adj}}$.

@@

Although each of the three model criteria suggested a different number of predictor variables, they were all within the same region (between 6 and 8 predictors). All three produced a similar error rate, so Occam's Razor was implemented and the simplest, six-variable model was selected. This was the model favored by BIC, which gives the greatest penalty for complexity in models. The six variables used in the regression context are `volatile.acidity`, `chlorides`, `total.sulfur.dioxide`, `pH`, `sulphates`, and `alcohol`. All of the variables from the classification selection appear with the exception of density. In the regression context, `pH` was the variable that proved to be a rather unintuitive inclusion, because it is similar to density in that it is not easily detected by human senses. The inclusion of these two variables in the model portray the importance of the chemical properties of a bottle of wine when it comes to quality.

@@

# Obtain RMSE and results for linear model
lm.reg.trn.pred <- predict(lm.reg, newdata = trn)
lm.reg.trn.rmse <- sqrt(mean((lm.reg.trn.pred - trn$quality)^2))

@@

Residuals indicate that the MLR assumptions are mostly satisfied. The residuals vs. fitted values plot (top left) has unusual stratified lines, but this effect is caused by the observed responses being integer values of 4, 5, 6, 7, etc. rather than the continuous values typical of a regression. The plot does not show heteroscedasticity and residuals are spread evenly around 0 over the entire range of fitted values. When residuals are standardized, as in the top right plot, one would expect about 95% to be inside the [-1.96, 1.96] interval; the actual percentage is `r round(100*scale.res.2std, 1)`%. The QQ-plot in the bottom left compares the normality of residuals against a theoretical distribution. Except for a slight downward bias of the lowest 5% of residuals at 2 standard deviations below 0, the data fits a normal distribution almost perfectly. On the bottom right plot, high leverage points (those with leverage $\ge$ 0.03) are not dramatically biased, being spread both above and below the 0 residual. Cook's Distance tops out at about 0.08, well below the generally accepted threshold of 1 that is considered highly influential. These plots taken together suggest that the data generally fulfills the requirements for modeling using MLR.

The largest benefit to using a multiple linear regression model is its interpretability and ease of use. The coefficients can be computed efficiently and they have clear and easy interpretations. It is also possible to make inference on any of the variables or linear combination of variables. The most significant drawback of using the multiple linear regression model is that the model is subject to distributional assumptions, meaning that if the assumptions are violated, the estimates and inference will be unstable at best and invalid at worst. But based on the residual plots it appears that the model does not violate any of the model assumptions.

@@

Based on cross validation, the random forest model for regression using two randomly-selected variables considered at each split was selected. This model produced a test error rate of `r round(rf.reg.test_err,4)`.

@@

In the regression setting, the test root mean squared error (RMSE) for both models were relatively low. The RMSE for the random forest was `r df.reg[2,2]` and the RMSE for the multiple linear regression model was `r df.reg[1,2]`, further supporting the hypothesis that the random forest would outperform the other models.

@@

Based on the results of the study, prediction of if a given wine has high or low quality was achieved with about 70-80% accuracy. Despite not being able to achieve 80% prediction accuracy with the basic classification tree, the random forest model showed an improvement and predicted with a higher accuracy. Although this is a *good* result considering the no-information rate of 53%, it is by no means excellent. These results indicate that machine learning algorithms could be a good technique when trying to predict wine quality, but that there is room for improvement. With a more formal, industry-expert-guided variable selection process, the team believes that higher prediction accuracy can be achieved. Machine learning models could potentially be combined with other techniques to add value to winemakers.
